---
title: "King County, Washington House Prices"
author: "Ian Knightly"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE, message=FALSE}
library(readr)
library(plotly)
library(ggplot)
library(tidyverse)
library(ggridges)
library(gridExtra)
library(dplyr)
library(car)

Sys.setenv("MAPBOX_TOKEN" = "pk.eyJ1IjoiaWtuaWdodGx5IiwiYSI6ImNsdWtkOGFqejBodHUyanFwcDdpb3p3d3MifQ.205T80MVLoDUpEPi_2kTDQ")

data<-read_csv("Preprocessed_Housing.csv")


```


## Introduction

- Data set obtained from (Kaggle.com).

- A comprehensive collection of property listings, encompassing various attributes such as the number of bedrooms, bathrooms, living area size, lot size, and location details in King County, WA.  

-  Aids in identifying trends and patterns in the real estate market, assisting investors, real estate agents, and policymakers in making informed decisions. 

- Also, serves as a foundation for developing recommendation systems for homebuyers, guiding them towards properties that align with their preferences and requirements. 


## Research Question and Hypothesis

-   **Research Question:** Does the ratio of bathrooms to bedrooms impact the price of a house?

-   $H_0$ The mean price of houses with different bathroom to bedroom ratios are the same i.e. $\mu_1 = \mu_2$

-   $H_a$ The mean price of houses with different bathroom to bedroom ratios are not the same i.e. $\mu_1 \neq \mu_2$.


## Exploratory Data Analysis (EDA)


- Log scaled price variable to ensure linearity among variables.

- Added variables
  - Bathroom-to-Bedroom Ratio: $\frac{bathroom}{bedroom}$.
  - Zip Code Region: Split King County into two regions w/ zip codes 980 and 981. 
  - Factor Levels: $\frac{bathroom}{bedroom} < 1$ and $\frac{bathroom}{bedroom} \geq 1$.
  - Blocks: Split the distribution of the total square feet variable into 4 blocks corresponding to quantile.
  
- Outliers for bathroom, bedroom, and total square feet were normalized to fall within 3 standard deviations.
- A Randomized Complete Block Design was performed on the data to test the difference in means between bath:bed.


## EDA Continued

```{r, echo=FALSE}
# Histogram of Bedrooms
plot_bedrooms <- plot_ly(data, x = ~bedrooms, type = 'histogram',
                         name = "Bedrooms", marker = list(color = 'lightblue')) %>%
                 layout(title = "Histogram of Bedrooms",
                        xaxis = list(title = "Number of Bedrooms"),
                        yaxis = list(title = "Count"))

# Histogram of Bathrooms
plot_bathrooms <- plot_ly(data, x = ~bathrooms, type = 'histogram',
                          name = "Bathrooms", marker = list(color = 'mediumblue')) %>%
                  layout(title = "Histogram of Bathrooms",
                         xaxis = list(title = "Number of Bathrooms"),
                         yaxis = list(title = "Count"))

# Histogram of Total Square Feet (Living Area)
plot_sqft_living <- plot_ly(data, x = ~sqft_living, type = 'histogram',
                            name = "Sqft Living", marker = list(color = 'darkblue')) %>%
                    layout(title = "Histogram of Total Square Feet (Living Area)",
                           xaxis = list(title = "Total Square Feet"),
                           yaxis = list(title = "Count"))

# Combine the histograms into a single subplot
combined_hist_plot <- subplot(plot_bedrooms, plot_bathrooms, plot_sqft_living, nrows = 3, shareX = FALSE, shareY = FALSE) %>%
                      layout(title = "Histograms of Bedrooms, Bathrooms, and Total Square Feet")
combined_hist_plot
```

## King County Map of Houses by Log Price

```{r, echo=FALSE}
#King County Map 
plot_mapbox() %>% add_markers(data = data, x = ~long, y = ~lat,
                              color = ~price,
                              alpha = I(0.9),
                              text = ~paste("The Log Price For This House is:",
                                            paste0("$",round(price,3))),
                              hoverinfo = "text") %>%
  layout(mapbox = list(
    center = list(lat = 47.53005, lon = -122.1139),
    zoom = 9))

```

## Ridgeline Plot

```{r, echo=FALSE, message=FALSE}
#Ridgeline Plot                       
#By Zip Code Region
data %>%
  ggplot(aes(x = price, y = `factor levels`,
             fill = `factor levels`)) +
  geom_density_ridges(alpha = 0.7) +
  stat_density_ridges(quantile_lines = TRUE, alpha = 0.75) +
  scale_fill_manual(values = c("lightblue", "darkblue")) +
  theme_ridges(font_size = 11) +
  labs(x = "Log Scale Price", fill = "Bath:Bed", y = "", title = "Ridgeline Plot of Log Price by Zip Code Region") +
  scale_x_discrete(labels = c("Fewer Bath:Bed", "More Bath:Bed")) +
  facet_wrap(vars(`zip code region`)) +
  theme(axis.title.x = element_text(hjust = 0.5),
        legend.key.size = unit(0.2, "cm"),
        legend.title=element_text(size=10), 
        legend.text=element_text(size=9))
```

## Shiny App

<https://iasios.shinyapps.io/633_final_app/>

## Box Plot

```{r, echo=FALSE}
boxplot<-ggplot(data, aes(x=`factor levels`, y=price)) +
  geom_boxplot(aes(fill = blocks)) +
  scale_fill_manual(values = c("lightblue",  "steelblue", "mediumblue", "darkblue")) +
  scale_x_discrete(labels = c("Fewer Bath:Bed", "More Bath:Bed")) +
  labs(
    title="Boxplots of Bathroom to Bedroom Ratio",
    x="Factor Levels",
    y="Log Scale Price of a House",
       fill = "Blocks") +
  theme_minimal() 

boxplot
```

## Scatter Plot

```{r, echo=FALSE}
scatterplot<-ggplot(data, aes(x=`factor levels`, y=price)) +
  geom_point(aes(fill=blocks, color = bed_bth_ratio)) +
  stat_summary(fun=mean, geom="point", shape=20, color="red", size=3) +
  stat_summary(fun=mean, geom="text", aes(label=round(after_stat(y), digits=1)), vjust=0.5, hjust = -0.5, color="mediumblue") +
  scale_x_discrete(labels = c("Fewer Bath:Bed", "More Bath:Bed")) +
  labs(
    title="Scatterplots of Bathroom to Bedroom Ratio",
    x="Factor Levels",
    y="Log Scale Price of a House",
       color = "Bath:Bed") +
  theme_minimal() + 
  scale_fill_discrete(guide = "none") +
  facet_wrap(~blocks)

scatterplot
```

## ANOVA Test

```{r, echo=FALSE}
anova_model<-aov(price~`factor levels` + blocks, data = data)
anova_summary<-summary(anova_model)
anova_components<-anova_summary[[1]]
anova_df<-as.data.frame(anova_components)
knitr::kable(anova_df, caption = "ANOVA Summary")
```

<div style="display: flex: 40%; justify-content: space-between;">
<div>

- $\alpha$ = 0.05
- P-Value < 0.0001
- $F^*$ = 270.4772
- F(0.95, 1, 3) = 10.1
- $F^*$ > F and P-Value < $\alpha$
- Conclude $H_a$

</div>
<div style="width: 75%;">

## Bootstrap

```{r, echo=FALSE}
# Bootstrap for Blocks 1-4)
# Function to extract data for a given block and factor level
extract_data <- function(block, factor_level) {
  data %>%
    filter(blocks == block, `factor levels` == factor_level) %>%
    pull(price)
}

# Function to perform bootstrapping and calculate confidence intervals
bootstrap_diff_means <- function(factor1_data, factor2_data, n = 1000) {
  diff_means <- numeric(n)
  
  for (i in 1:n) {
    fctr1_sample <- sample(factor1_data, length(factor1_data), replace = TRUE)
    fctr2_sample <- sample(factor2_data, length(factor2_data), replace = TRUE)
    fctr1_btstrp<-factor1_data[fctr1_sample]
    fctr2_btstrp<-factor2_data[fctr2_sample]
    diff_means[i] <- mean(fctr2_btstrp, na.rm = TRUE) - mean(fctr1_btstrp, na.rm = TRUE)
  }
  
  ci <- quantile(diff_means, c(0.05, 0.95), na.rm = TRUE)
  list(diff_means = diff_means, ci = ci)
}

# List of blocks and factor levels
blocks <- c("Block 1", "Block 2", "Block 3", "Block 4")
factor_levels <- c("Factor Level 1", "Factor Level 2")

# Iterate through blocks to calculate original differences and bootstrap confidence intervals
results <- lapply(blocks, function(block) {
  factor1_data <- extract_data(block, "Factor Level 1")
  factor2_data <- extract_data(block, "Factor Level 2")
  

  original_diff <- mean(factor2_data) - mean(factor1_data)
  bootstrap_results <- bootstrap_diff_means(factor1_data, factor2_data)
 

  list(
    block = block,
    original_diff = original_diff,
    confidence_interval = bootstrap_results$ci
  )
})

# Print results
for (res in results) {
  cat(paste0(res$block, " Original Difference: ", res$original_diff, "\n"))
  cat(paste0(res$block, " 95% Confidence Interval: [", res$confidence_interval[1], ", ", res$confidence_interval[2], "]\n"))
}
```

```{r, echo=FALSE}
btstrp_fctr1<-data%>%filter(`factor levels` == "Factor Level 1")%>%select(price)%>%pull(price)
btstrp_fctr2<-data%>%filter(`factor levels` == "Factor Level 2")%>%select(price)%>%pull(price)


original_diff<-mean(btstrp_fctr2) - mean(btstrp_fctr1)


# Initialize a vector to store the differences
diff_means <- numeric(1000)



# Bootstrap
for(i in 1:1000){
  # Create a bootstrap sample
  # Sample with replacement from the indices of the data
  # This will create a new sample of the same size as the original data
  fctr1_bootstrap_sample <- sample(length(btstrp_fctr1), replace = TRUE) 
  fctr2_bootstrap_sample <- sample(length(btstrp_fctr2), replace = TRUE) 
  
 
  # Create the bootstrap samples
  # Use the sampled indices to create the bootstrap samples
  fctr1_btstrp<- btstrp_fctr1[fctr1_bootstrap_sample]
  fctr2_btstrp<- btstrp_fctr2[fctr2_bootstrap_sample]

  # Store the difference in means
  diff_means[i] <- mean(fctr2_btstrp) - mean(fctr1_btstrp)
 
}

# Calculate the Block 1 confidence interval
ci <- quantile(diff_means, c(0.05, 0.95))
print(paste("Block 1 Original Difference: ", block1_original_diff))
print(paste("Block 1 95% Confidence Interval: [", block1_ci[1], ", ", block1_ci[2], "]"))



histogram1<-ggplot(data.frame(x = diff_means)) + 
  geom_histogram(aes(x = x), color = "white", bins = 40) + 
  geom_vline(xintercept = ci[1], color = "blue") +
  geom_vline(xintercept = ci[2], color = "blue") +
  geom_vline(xintercept = mean(diff_means), color = "black") +
  geom_vline(xintercept = original_diff, color = "red") +
  theme_minimal() + 
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) + ggtitle("Bootstrap Histogram") +
  xlab("Difference in Means") + ylab("Density") +
  scale_x_continuous(breaks = seq(-0.2, 0.2, by = 0.01)) +
  scale_y_continuous(breaks = seq(0, 50, by = 5)) +
  geom_text(aes(x = 0.109, y = 50, label = "Lower CI"), vjust = -1) +
  geom_text(aes(x = 0.1185, y = 50, label = "Upper CI"), vjust = -1) +
  geom_text(aes(x = 0.1185, y = 40, label = "Original Difference"), vjust = -1) +
  labs(title = "Block 1 Bootstrap w/ 95% CI", x = "Difference in Means", y = "Density")

histogram1
```


## Conclusion

- There is significant difference between house prices concerning the number of bathrooms each house has.

- Adding an extra half-bathroom could potentially increase the value of your home.

- Construction contractors can net a greater profit by building homes with an additional bathroom. 

- Real Estate agents can increase their commissions by selling homes with an additional bathroom.


## References

<https://www.kaggle.com/datasets/sukhmandeepsinghbrar/housing-price-dataset>
<https://www.mapbox.com/>
